---
title: "Class05: Data Viz with GGPLOT"
author: "Jada Brown (PID: A15573784)"
format: pdf
---

Today we are playing with plotting and graphics in R.

There are lots of ways to make cool figures in R.
There are "base" R graphics (`plot()`, `hist()`, `boxplot()`, etc,)

There are also add-on packages like **ggplot**.

```{r}
head(cars, 3)
```
Let's plot this with "base" R:

```{r}
plot(cars)
```

```{r}
head(mtcars)
```

Let's plot `mpg` vs `disp`
```{r}
plot(mtcars$mpg,mtcars$disp, pch = 17)
```

```{r}
hist(mtcars$mpg)
```

##GGPLOT 

The main function in the ggplot2 package is **ggplot()**
First I need to install the **ggplot2** pacjage. I can install any package with the function, `install.packages("")`
> **N.B** I never want to run `install packages("")` in my quarto document!!!

```{r}
library(ggplot2)
ggplot(cars) +
  aes(x = speed, y = dist) +
  geom_point()
```

```{r}
ggplot(cars) +
  aes(speed) +
  geom_histogram()
```

Every ggplot needs at least 3 things: 

- The **data** (given with `ggplot(cars)`) +
- The **aesthetic** mapping (given with `aes()`) +
- The **geom_** given by `geom_point()`)

> For simple canned graphs, "base" R is nearly always faster

###Adding more layers

Let's add a line and a title, subtitle and caption as well as custom axis labels

```{r}
ggplot(cars) +
  aes(x = speed, y = dist) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  labs(title = "Silly Plot", subtitle = "Speed v Distance", x = "Speed (MPH)", y = "Distance (ft)", caption = "when is tea time anyway???") +
  theme_bw()
```

## Plot some expression data

Read data file from online URL

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```
> Q1. How many genes are in this wee data?

There are `r nrow(genes)` in this dataset

```{r}
nrow(genes)
ncol(genes)
```

> Q2. How many "up" regulated genes are there?

```{r}
sum(genes$State == "up")
```

OR

```{r}
table (genes$State)
```

There are `r sum(genes$State == "up")` up genes 



```{r}
p <- ggplot(genes) +
  aes(x = Condition1, y = Condition2, col = State) +
  geom_point() +
  scale_color_manual(values = c("blue", "black", "red"))
```

```{r}
p + labs(title="look at me")
```

Silly example of adding labels

```{r}
library(ggrepel)
```


```{r}
ggplot(genes) +
  aes(x = Condition1, 
      y = Condition2, 
      col = State,
      label = Gene) +
  geom_point() +
  scale_color_manual(values = c("blue", "grey", "red")) +
  geom_text_repel(max.overlaps = 60) +
  theme_bw()
```


## Going Further

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```

```{r}
head(gapminder)
```

```{r}
tail(gapminder, 6)
```

A first plot

```{r}
ggplot(gapminder) +
  aes(y = lifeExp, x = gdpPercap, col = continent) +
  geom_point() +
  facet_wrap(~continent) +
  theme_bw()
```

```{r}
p2 <- ggplot(gapminder) +
  aes(y = lifeExp, x = gdpPercap, col = continent, size = pop) +
  geom_point(alpha=0.5) +
  facet_wrap(~continent) +
  theme_bw() +
  scale_size_area(max_size = 10)
```

```{r}
library(patchwork)
```

```{r}
(p | p2)
```

