---
title: "Lab10: Halloween Special"
author: "Jada Brown (PID: A15573784)"
format: pdf
---

```{r}
candy_file <- read.csv("./candy-data.csv")
head(candy_file)
```

```{r}
library(flextable)
flextable::flextable( head(candy_file))
```

```{r}
candy <- read.csv("./candy-data.csv", row.names = 1)
head(candy_file)
```


> Q1. How many different candy types are in this dataset?

There are 85 types of candy in this dataset.
```{r}
nrow(candy)
```
```{r, message = FALSE}
library(dplyr)

candy |> nrow()
```


> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

%>% or |>


> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Snickers",]$winpercent
```


> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```


> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

```{r}
library(skimr)
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

 yes winpercent
> Q7. What do you think a zero and one represent for the candy$chocolate column?

Its a yes or no that the candy in question is chocolate.

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)

ggplot(candy) + aes(winpercent) + geom_histogram(bins=20)
```

> Q9. Is the distribution of winpercent values symmetrical?
 
 No
 
```{r}
ggplot(candy) + aes(winpercent) + geom_density()
```
 
 
> Q10. Is the center of the distribution above or below 50%?

```{r}
mean(candy$winpercent)
```

```{r}
summary(candy$winpercent)
```


> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

Yes, the mean is higher in chocolate candy.

```{r}
#1. Find all chocolate candy in dataset
choco <- as.logical(candy$chocolate)
choco_candy <- candy[choco,]
choco_candy
#2. Extract their `winpercent` values
chocowin <- choco_candy$winpercent
chocowin
#3. Find the mean of these values
mean_choc <- mean(chocowin)
#4-6. Repeat for Fruit candy
fruity <- as.logical(candy$fruity)
fruitcandy <- candy[fruity,]
fruitcandy
fruitwin <-fruitcandy$winpercent
fruitwin
mean_fruit <- mean(fruitwin)
#7. Which mean value is higher?
mean_fruit 
mean_choc
```

> Q12. Is this difference statistically significant?

Yes, the p-value is statistically significant.

```{r}
t.test(chocowin, fruitwin)
```

> Q13. What are the five least liked candy types in this set?

Nik L Nip, Boston Baked, Chicklets, Super Bubble, Super Bubble, Jawbusters
```{r}
head(candy[order(candy$winpercent),], n=5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
tail(candy[order(candy$winpercent),], n=5)
```

Snickers, Kit Kat, Twix, Reese's Miniatures, Reese's Peanut Butter cup

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy) + aes(winpercent, rownames(candy)) + geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + aes(winpercent, reorder(rownames(candy), winpercent)) + geom_col()
```
Add some color:
```{r}
mycols <- rep("black", nrow(candy))
mycols[candy$chocolate==1] <- "chocolate"
mycols[candy$bar==1] <- "brown"
mycols[candy$fruity==1] <- "pink"
mycols
  
ggplot(candy) + aes(x = winpercent, y = reorder(rownames(candy), winpercent)) + geom_col(fill = mycols)
```

> Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is Sixlets

> Q18. What is the best ranked fruity candy?

The best ranked fruity candy is Starbursts

## Winpercent v Pricepercent

```{r}
library(ggrepel)

mycols[candy$fruity==1] <- "red"
mycols[candy$chocolate==1] <- "chocolate"
mycols[candy$bar==1] <- "brown"

ggplot(candy) + aes(winpercent, pricepercent, label=row.names(candy)) + geom_point(col=mycols) + geom_text_repel(col=mycols)
```
> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

The candy that is the highest rank in terms of winpercent for the least money is Reese's Miniatures.


> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
best <- order(candy$pricepercent, decreasing = TRUE)
head( candy[best,c(11,12)], n=5 )
```

Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, Hershey's Milk Chocolate are the most expensive.

## Correlation

```{r}
cij <- cor(candy)
```

```{r}
library(corrplot)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruit, and chocolate are strongly anti-correlated.

> Q23. Similarly, what two variables are most positively correlated?

The most positively correlated variables are the ones that are the same, ex. Chocolate and chocolate. Aside from that, the next two variables are chocolate and winpercent.

## Principal Component Analysis

The main function in base R for this is `prcomp()` and we want to `scale=TRUE` here:

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

Let's look at our first main result figure - the "PC Plot" or PC1 v PC2:

```{r}
ggplot(pca$x) + aes(PC1, PC2, label = row.names(pca$x)) + geom_point(col=mycols) + geom_text_repel(col = mycols) + labs(title = "Jada")
```
Dont forget about your variable loadings - how the original variables contribute to your new PCs...

```{r}
ggplot(pca$rotation) + aes(PC1, row.names(pca$rotation)) + geom_col()
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

The variables strongly picked up by PC1 in the positive directions are Pluribus, hard, and fruity.
