---
title: "HW Class 6"
author: Jada Brown (PID: A15573784)
format: pdf
---

```{r}
library(bio3d)
s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug

s1
```

## Q1. What type of object is returned from the read.pdb() function? 
A variable that 


```{r}
s1.chainA <- trim.pdb(s1, chain="A", elety="CA")
s2.chainA <- trim.pdb(s2, chain="A", elety="CA")
s3.chainA <- trim.pdb(s1, chain="A", elety="CA")

s1.chainA
```

## Q2. What does the trim.pdb() function do?
The `trim.pdb()` function trims the data by showing a information about the specific domain of the protein of interest.

```{r}
s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b

s1.b
```
```{r}
plotb3(s1.b, s1.chainA, typ="l", ylab="Bfactor")
plotb3(s2.b, sse=s2.chainA, typ="l", ylab="Bfactor")
plotb3(s3.b, sse=s3.chainA, typ="l", ylab="Bfactor")
```

## Q3. What input parameter would turn off the marginal black and grey rectangles in the plots and what do they represent in this case? 
Remove the Marginal black and grey rectangles in the plots by deleting "sse

## Q4. What would be a better plot to compare across the different proteins?



## Q5. Which proteins are more similar to each other in their B-factor trends. How could you quantify this? 



